% \begin{savequote}[8cm]
%   \textlatin{Cor animalium, fundamentum e\longs t vitæ, princeps omnium, Microco\longs mi Sol, a quo omnis vegetatio dependet, vigor omnis \& robur emanat.}

%   The heart of animals is the foundation of their life, the sovereign of everything within them, the sun of their microcosm, that upon which all growth depends, from which all power proceeds.
%   \qauthor{--- William Harvey \cite{harvey_exercitatio_1628}}
% \end{savequote}

\chapter{\label{app:background}Appendix \textemdash\ Background}
\minitoc

\section{Background Literature Search}

A literature search was performed with the aim of writing a narrative review on the VL relapse. Specifically to address the following questions:

\begin{itemize}[noitemsep]
  \item What is the burden of relapse?
  \item When does relapse occur?
  \item What is the cause (mechanism) of relapse?
  \item What are the determinants of relapse?
\end{itemize}

All searches performed on August 11\textsuperscript{th} 2025 on PubMed, Embase and Web of Science (Boxes A.1--A.3). Search performed from database inception. After deduplication with Covidence\cite{covidence} 1,891 articles were identified. Following title and abstract review, 55 articles were identified for inclusion in the narrative review.

Inclusion criteria:
\begin{description}
  \item[Inclusion criteria:](1) Publications looking at associations between VL relapse occurrence and host/parasite characteristics during, before, or shortly after treatment of index case. (2) Systematic/literature reviews looking at relapse determinants. (3) All ages, geographical locations.
  \item[Exclusion criteria:](1) Conference abstracts, protocols, case reports. (2) Articles not published in English. (3) Non-human studies.
\end{description}

\begin{mybox}[label=box:search1]{PubMed search terms (932 hits)}
  \texttt{(``Leishmaniasis, Visceral''[Mesh] OR ``visceral leishmaniasis'' OR ``leishmaniasis, visceral'' OR ``kala azar'' OR ``kala--azar'')
  AND
  (``Recurrence''[Mesh] OR relapse* OR recurrent OR recurrence OR recrudescence OR ``treatment failure'')}
\end{mybox}

\begin{mybox}[label=box:search2]{Embase search terms (1303 hits)}
  {\ttfamily
    \begin{enumerate}[noitemsep]
      \item visceral leishmaniasis.mp. or exp visceral leishmaniasis/
      \item kala--azar.mp.
      \item 1 or 2
      \item exp relapse/ or relapse*.mp.
      \item recurrence*.mp. or exp recurrent disease/
      \item recurrent.mp.
      \item recrudescence.mp.
      \item 4 or 5 or 6 or 7
      \item treatment failure.mp. or exp treatment failure/
      \item 8 or 9
      \item 3 and 10
    \end{enumerate}
  }
\end{mybox}

\begin{mybox}[label=box:search3]{Web of Science (1097 hits)}
  {\ttfamily
    \begin{enumerate}[noitemsep]
      \item TS=((Leishmaniasis and Visceral) or (Kala azar) or (Kala azar))
      \item TS=((relapse\$) or (recurren\$) or (recrudescence) or  (treatment failure\$))
      \item \#2 AND \#1
    \end{enumerate}
  }
\end{mybox}

